Consistent ecosystem functional response during five years of drought and deluge in a sagebrush steppe

### 2017-07-11 manuscript/usses_water_ms.Rmd
finished discussion draft; ready for comments

---
title: Consistent ecosystem functional response during five years of drought and deluge in a sagebrush steppe
preprint: false
author: 
  - name: Andrew T. Tredennick
    affiliation: 1
    corresponding: true
    email: atredenn@gmail.com
  - name: Andrew R. Kleinhesselink
    affiliation: 2
  - name: Peter B. Adler
    affiliation: 1
affiliation:
  - code: 1
    address: Department of Wildland Resources and the Ecology Center, Utah State University, Logan, Utah 84322
  - code: 2
    address: Department of Ecology and Evolutionary Biology, University of California, Los Angeles, Los Angeles, California xxxxx
abstract: >
  Precipitation is predicted to become more variable in the western U.S., meaning years of above and below average precipitation will become more common. Such periods of drought and deluge could become major drivers of plant community dynamics and ecosystem functioning in water limited grasslands. Here we report the results of a five-year experiment where we used drought and irrigation treatments (50% decrease/increase) to see how a sagebrush steppe plant community in Idaho will respond to future climate changes. The ecosystem was suprisingly resistant to both drought and irrigation. Aboveground net primary productivity (ANPP) responded positively to ambient growing season precipitation, but the response did not vary across treatments. There was also no evidence that treatment effects grew over time. The similarity of ecosystem functioning was not due to compensatory shifts at the plant community level, where species composition among treatments was similar and remarkably stable over the five years. At least in the short-term, ecosystem functioning and community composition in this sagebrush steppe system is resistant to increases and decreases in growing season precipitation.
header-includes: >
  \usepackage{lipsum}
  \usepackage{setspace}
  \usepackage{todonotes}
  \usepackage{rotating}
  \usepackage{color, soul}
  \usepackage{sectsty}
  \usepackage{bm,mathrsfs}
bibliography: ~/Dropbox/Bibliography/usses_water.bib
output: 
  rticles::peerj_article:
    base_format: rmarkdown::pdf_document # bookdown::pdf_document2 # for using \@ref()
---

@@

@@

\noindent{} where $\mu_{i(j(k))}$ is the deterministic prediction from the regression model for observation *i* for plot *j* associated with treatment *k* in year *t*, $\beta_{0,j(k)}$ is the intercept for plot *j* associated with treatment *k*, $\beta_{1,j(k)}$ is the slope term for the effect of soil moisture for plot *j* associated with treatment *k*, $\gamma_t$ is the intercept offset for year *t*, and $\sigma^2_k$ is the process variance for treatment *k*.
Data include the standardized log(ANPP) observations $\left( y_{i(j(k(t)))} \right)$ and soil moisture ($x_i$).
Although we include observation subscript *i* on the *x*s, observations within a treatment-year all share the same soil moisture values.

@@

@@

\noindent{} where $\boldsymbol{\beta}_{j(k)}$ is the vector of regression coefficients (intercept and slope) for plot *j* associated with treatment *k*, $\boldsymbol{\beta}_{k}$ is the vector of coefficients for each treatment, and $\boldsymbol{\beta}$ is the vector of overall coefficients.
The plot- and treatment-level coefficients are drawn from multivariate normal distributions with covariance matrix $\boldsymbol{\Sigma}$.
For the plot-level coefficients, each treatment has its own variance-covariance matrix (i.e., $\boldsymbol{\Sigma}(k)$).
The overall coefficients are drawn from a normal prior with mean 0 and standard deviation 1.
The random year effects ($\boldsymbol{\gamma}$) are drawn from a normal prior with mean 0 and standard deviation $\sigma_{\text{year}}$, which was drawn from a weibull distribution.
A full description of model is in Appendix 2.

@@

Three of our five treatment years fell in years of below average rainfall (Fig. 1A).
Thus, those three years represent a lower magnitude of absolute change in precipitation experienced by the treatments.
Averaged across treatments, ANPP varied from a minimum of 74.5 g m$^{-2}$ in 2014 to a maximum of 237.1 g m$^{-2}$ in 2016 (Fig. 1C).
ANPP was slightly higher in irrigation plots and slightly lower in drought plots (Fig. 1C), corresponding to estimated soil volumetric water content (VWC) differences among treatments (Fig. 1B).
Such differences in soil VWC indicate our treatment infrastructure was successful.

Cumulative growing season soil moisture had a positive effect on ANPP (mean of $\beta_{1}$ = 0.43; 80\% BCI = -0.05, 0.88; 95\% BCI = -0.36, 1.15) (Fig. 1D).
Average ANPP was similar among treatments (similar intercepts, Fig. 2A), as was the effect of precipitation (similar slopes, Fig. 2B).
In an average precipitation year (i.e., $x$ = 0 in Eq. 1), the probability that ANPP in a conrol plot is greater than ANPP in a drought plot was 0.71, and the probability that ANPP in a control plot is lower than in an irrigation plot was 0.41.
In other words, the posterior distributions of $\beta_0(\text{control}) - \beta_0(\text{drought})$ and $\beta_0(\text{control}) - \beta_0(\text{drought})$ broadly overlapped zero.
There was also no evidence that the treatment effects became more important over time because there was no directional trend in the yearly treatment effects estimated from the ANPP-Treatment model (Fig. 3).

Community composition was similar among treatments.
In no year did community composition among treatments not overlap, and they were equally dispersed in all years (Table 1; Fig. 4).
Likewise, community composition was remarkably stable over time, with no evidence of divergence among treatments (Table 1; Fig. 4).
Species' abundances and ranks showed little deviation over the five-year experiment, regardless of treatment (Fig. 5).

@@

Surprisingly, ANPP was not sensitive to either drought or irrigation treatments, whether in the context of growing season precipitation (Fig. 2) or in the context of year since the experiment began (Fig. 3).
This is surprising because grasslands generally, and sagebrush steppe specifically, are considered water-limited systems.
Thus, our expectation was that ANPP would be tightly linked to soil water availability.
ANPP was slightly lower in drought plots and slightly higher in irrigation plots, but these differences were not statistically important.

In the absence of community composition data, an obvious hypothesis to explain the lack of ANPP response is that species reordered rapidly.
In so doing, newly abundant species could compensate for the loss of ecosystem functioning of previously abundant species.
We did not find this to be the case, as specie composition was remarkably stable over the course of the experiment (Fig. 3).

In combination, the lack of ecosystem and community response to drought or irrigation shows that this sagebrush steppe ecosystem is resistant to chronic alterations in water availability.
There are three possible explanations for the resistance we found.
First, it could be that our experiment simply was not long enough to induce responses.
This may be true for community responses, which can take decades [@Wilcox2016], but the response of grassland ANPP to water manipulations is typically immediate [e.g., @Hoover2014].

Second, it could be that our manipulations were not large enough to induce a response.
That is, maybe a 50\% decrease/increase in any given year is not abnormal give our sites historical range of variability [@Knapp2017].
We cannot definitively rule out this possibility, but we have reason to believe our manipulations *should* have been large enough.
Using the methods described by @Lemoine2016, we calculated the percent reduction and increase of mean growing season precipitation neccesary to reach the 1\% and 99\% extremes of the historical precipitation regime at our site.
The 1\% quantile of precipitation at our site is 110 mm, a 47% reduction from the mean, and the 99\% quantile is 414 mm, a 77\% increase from mean growing season precipitation (Appendix 3).
Thus, our drought treatment represented extreme precipitation amounts, especially in years where ambient precipitation was below average (Fig. 1A).
The irrigation treatment may have been too small, however.

Third, the ecosystem and community resistance to drought and deluge we observed could be a real phenomenon in this system.

>>>>

---
title: Consistent ecosystem functional response during five years of drought and deluge in a sagebrush steppe
preprint: false
author: 
  - name: Andrew T. Tredennick
    affiliation: 1
    corresponding: true
    email: atredenn@gmail.com
  - name: Andrew R. Kleinhesselink
    affiliation: 2
  - name: Peter B. Adler
    affiliation: 1
affiliation:
  - code: 1
    address: Department of Wildland Resources and the Ecology Center, Utah State University, Logan, Utah 84322
  - code: 2
    address: Department of Ecology and Evolutionary Biology, University of California, Los Angeles, Los Angeles, California 90095
abstract: >
  Precipitation is predicted to become more variable in the western U.S., meaning years of above and below average precipitation will become more common. Such periods of drought and deluge could become major drivers of plant community dynamics and ecosystem functioning in water limited grasslands. Here we report the results of a five-year experiment where we used drought and irrigation treatments (50% decrease/increase) to see how a sagebrush steppe plant community in Idaho will respond to future climate changes. Ecosystem functional response, defined as the relationship between soil moisture and aboveground net primary productivity, was suprisingly consistent across drought, control, and irrigation treatments. Aboveground net primary productivity (ANPP) responded positively to availabile soil moisture, but the response did not vary across treatments. There was also no evidence that treatment effects grew over time. The similarity of ecosystem functional response across treatment was not due to compensatory shifts at the plant community level, where species composition among treatments was similar and remarkably stable over the five years. At least in the short-term, ecosystem functional response and community composition in this sagebrush steppe system is resistant to increases and decreases in growing season precipitation.
header-includes: >
  \usepackage{lipsum}
  \usepackage{setspace}
  \usepackage{todonotes}
  \usepackage{rotating}
  \usepackage{color, soul}
  \usepackage{sectsty}
  \usepackage{bm,mathrsfs}
bibliography: ~/Dropbox/Bibliography/usses_water.bib
output: 
  rticles::peerj_article:
    base_format: rmarkdown::pdf_document # bookdown::pdf_document2 # for using \@ref()
---

@@

\ark{ANDY: add paragraph on the statistical modeling of soil moisture to produce Figure 2B (same as your Figure 2 in the precip-demography-prediction dissertation chapter).}

@@

\noindent{}where $\mu_{i(j(k))}$ is the deterministic prediction from the regression model for observation *i* for plot *j* associated with treatment *k* in year *t*, $\beta_{0,j(k)}$ is the intercept for plot *j* associated with treatment *k*, $\beta_{1,j(k)}$ is the slope term for the effect of soil moisture for plot *j* associated with treatment *k*, $\gamma_t$ is the intercept offset for year *t*, and $\sigma^2_k$ is the process variance for treatment *k*.
Data include the standardized log(ANPP) observations $\left( y_{i(j(k(t)))} \right)$ and soil moisture ($x_i$).
Although we include observation subscript *i* on the *x*s, observations within a treatment-year all share the same soil moisture values.

@@

\vspace{-1em}

@@

\noindent{}where $\boldsymbol{\beta}_{j(k)}$ is the vector of regression coefficients (intercept and slope) for plot *j* associated with treatment *k*, $\boldsymbol{\beta}_{k}$ is the vector of coefficients for each treatment, and $\boldsymbol{\beta}$ is the vector of overall coefficients.
The plot- and treatment-level coefficients are drawn from multivariate normal distributions with covariance matrix $\boldsymbol{\Sigma}$.
For the plot-level coefficients, each treatment has its own variance-covariance matrix (i.e., $\boldsymbol{\Sigma}(k)$).
The overall coefficients are drawn from a normal prior with mean 0 and standard deviation 1.
The random year effects ($\boldsymbol{\gamma}$) are drawn from a normal prior with mean 0 and standard deviation $\sigma_{\text{year}}$, which was drawn from a weibull distribution.
A full description of model is in Appendix 2.

@@

Three of our five treatment years fell in years of below average rainfall (Fig. 2A).
Thus, those three years represent a lower magnitude of absolute change in precipitation experienced by the treatments.
Averaged across treatments, ANPP varied from a minimum of 74.5 g m$^{-2}$ in 2014 to a maximum of 237.1 g m$^{-2}$ in 2016 (Fig. 2C).
ANPP was slightly higher in irrigation plots and slightly lower in drought plots (Fig. 2C), corresponding to estimated soil volumetric water content (VWC) differences among treatments (Fig. 2B).
Such differences in soil VWC indicate our treatment infrastructure was successful.

Cumulative growing season soil moisture had a positive effect on ANPP (mean of $\beta_{1}$ = 0.43; 80\% BCI = -0.05, 0.88; 95\% BCI = -0.36, 1.15) (Fig. 2D).
Ecosystem functional response was similar among treatments, with treatment level regressions between soil moisture and log(ANPP) having similar intercepts (Fig. 3A) and slopes (Fig. 3B).
There was also no evidence that the treatment effects became more important over time because there was no directional trend in the random year effects (Fig. 4).

Community composition was similar among treatments.
In no year did community composition among treatments not overlap, and they were equally dispersed in all years (Table 1; Fig. 5).
Likewise, community composition was remarkably stable over time, with no evidence of divergence among treatments (Table 1; Fig. 5).

@@

Ecosystem functional response (i.e., the relationship between soil moisture and ANPP) and community composition in this sagebrush steppe ecosystem was consistent across drought, control, and irrigation treatments (Fig. 2D, Fig. 3).
In other words, functional responses estimated from any precipitation level sampled apply to any other level of precipitation, and the consistent functional response was not due to species reordering and compensation.
Thus, our results conform to the first of our four predictions: neither ecosystem functional response nor community composition changed with chronic alteration in water availability (Fig. 1, top left).

This is surprising because grasslands generally, and sagebrush steppe specifically, are considered water-limited systems.
Thus, our naive expecatations were that either (i) ecosystem functional response would differ among treatments as species reach physiological thresholds and resource limitations (Fig. 1, top right), or (ii) ecosystem functional response stays the same due to species reordering and compensation (Fig. 1, bottom left).

In the absence of community composition data, an obvious hypothesis to explain the consistent ecosystem functional response is that species reordered rapidly.
In so doing, newly abundant species could compensate for the loss of ecosystem functioning of previously abundant species.
We did not find this to be the case, as species composition was remarkably stable over the course of the experiment (Fig. 5).

In combination, the lack of ecosystem and community response to drought or irrigation shows that this sagebrush steppe ecosystem is resistant to chronic alterations in water availability.
There are three possible explanations for the resistance we found.
First, it could be that our experiment simply was not long enough to induce responses.
This may be true for community responses, which can take decades [@Wilcox2016].

Second, it could be that our manipulations were not large enough to induce a response.
That is, maybe a 50\% decrease/increase in any given year is not abnormal give our sites historical range of variability [@Knapp2017].
We cannot definitively rule out this possibility, but we have reason to believe our manipulations *should* have been large enough.
Using the methods described by @Lemoine2016, we calculated the percent reduction and increase of mean growing season precipitation neccesary to reach the 1\% and 99\% extremes of the historical precipitation regime at our site.
The 1\% quantile of precipitation at our site is 110 mm, a 47% reduction from the mean, and the 99\% quantile is 414 mm, a 77\% increase from mean growing season precipitation (Appendix 3).
Thus, our drought treatment represented extreme precipitation amounts, especially in years where ambient precipitation was below average (Fig. 2A).
The irrigation treatment may have been too small, however.

Third, the ecosystem and community resistance to drought and deluge we observed could be a real phenomenon in this system, at least over the five year time span we observed.
The perennial plants in this cold desert ecosystem are tolerant to drought conditions [@Bazzaz1979; @Franks2011, A.R. Kleinhesselink, unpublished data], meaning a consistent ecosystem functional response at low soil moisture is perfectly reasonable.
For example, many of the perennial grasses in our focal ecosystem avoid drought stress by growing early in the growing season (A.R. Kleinhesselink, personal observation).
Likewise, the dominant shrub in our focal ecosystem, *Artemisia tripartita*, has access to water deep in the soil profile thanks to a deep root system [@Germino2014].
As for the irrigation treatments, the lack of a saturating response (e.g., a lower regression slope) suggests water availability remained a limiting factor at the levels of precipitation we were able to manipulate.

In conclusion, our results suggest that five years of $\pm$ 50\% ambient precipitation is not enough to induce a shift in ecosystem functional response in a sagebrush steppe.
This is likely because component species are tolerant to drought conditions and water remains co-limiting at higher precipitation levels.
Longer time series of chronic precipitation alteration may reveal plant community shifts that we did not observe [e.g., @Wilcox2016].
Our results suggest compositional shifts would have the largest impact at high rainfall because the current community maintained consistent ecosystem functional response at very low water availability.

### 2017-07-11 manuscript/usses_water_ms.tex
finished discussion draft; ready for comments

\affil[1]{Department of Wildland Resources and the Ecology Center, Utah State
University, Logan, Utah 84322}
\affil[2]{Department of Ecology and Evolutionary Biology, University of
California, Los Angeles, Los Angeles, California xxxxx}

@@

% 
\begin{abstract}
Precipitation is predicted to become more variable in the western U.S.,
meaning years of above and below average precipitation will become more
common. Such periods of drought and deluge could become major drivers of
plant community dynamics and ecosystem functioning in water limited
grasslands. Here we report the results of a five-year experiment where
we used drought and irrigation treatments (50\% decrease/increase) to
see how a sagebrush steppe plant community in Idaho will respond to
future climate changes. The ecosystem was suprisingly resistant to both
drought and irrigation. Aboveground net primary productivity (ANPP)
responded positively to ambient growing season precipitation, but the
response did not vary across treatments. There was also no evidence that
treatment effects grew over time. The similarity of ecosystem
functioning was not due to compensatory shifts at the plant community
level, where species composition among treatments was similar and
remarkably stable over the five years. At least in the short-term,
ecosystem functioning and community composition in this sagebrush steppe
system is resistant to increases and decreases in growing season
precipitation.
% Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text.
\end{abstract}

@@

\noindent{} where \(\mu_{i(j(k))}\) is the deterministic prediction from
the regression model for observation \emph{i} for plot \emph{j}
associated with treatment \emph{k} in year \emph{t}, \(\beta_{0,j(k)}\)
is the intercept for plot \emph{j} associated with treatment \emph{k},
\(\beta_{1,j(k)}\) is the slope term for the effect of soil moisture for
plot \emph{j} associated with treatment \emph{k}, \(\gamma_t\) is the
intercept offset for year \emph{t}, and \(\sigma^2_k\) is the process
variance for treatment \emph{k}. Data include the standardized log(ANPP)
observations \(\left( y_{i(j(k(t)))} \right)\) and soil moisture
(\(x_i\)). Although we include observation subscript \emph{i} on the
\emph{x}s, observations within a treatment-year all share the same soil
moisture values.

@@

\noindent{} where \(\boldsymbol{\beta}_{j(k)}\) is the vector of
regression coefficients (intercept and slope) for plot \emph{j}
associated with treatment \emph{k}, \(\boldsymbol{\beta}_{k}\) is the
vector of coefficients for each treatment, and \(\boldsymbol{\beta}\) is
the vector of overall coefficients. The plot- and treatment-level
coefficients are drawn from multivariate normal distributions with
covariance matrix \(\boldsymbol{\Sigma}\). For the plot-level
coefficients, each treatment has its own variance-covariance matrix
(i.e., \(\boldsymbol{\Sigma}(k)\)). The overall coefficients are drawn
from a normal prior with mean 0 and standard deviation 1. The random
year effects (\(\boldsymbol{\gamma}\)) are drawn from a normal prior
with mean 0 and standard deviation \(\sigma_{\text{year}}\), which was
drawn from a weibull distribution. A full description of model is in
Appendix 2.

@@

Three of our five treatment years fell in years of below average
rainfall (Fig. 1A). Thus, those three years represent a lower magnitude
of absolute change in precipitation experienced by the treatments.
Averaged across treatments, ANPP varied from a minimum of 74.5 g
m\(^{-2}\) in 2014 to a maximum of 237.1 g m\(^{-2}\) in 2016 (Fig. 1C).
ANPP was slightly higher in irrigation plots and slightly lower in
drought plots (Fig. 1C), corresponding to estimated soil volumetric
water content (VWC) differences among treatments (Fig. 1B). Such
differences in soil VWC indicate our treatment infrastructure was
successful.

Cumulative growing season soil moisture had a positive effect on ANPP
(mean of \(\beta_{1}\) = 0.43; 80\% BCI = -0.05, 0.88; 95\% BCI = -0.36,
1.15) (Fig. 1D). Average ANPP was similar among treatments (similar
intercepts, Fig. 2A), as was the effect of precipitation (similar
slopes, Fig. 2B). In an average precipitation year (i.e., \(x\) = 0 in
Eq. 1), the probability that ANPP in a conrol plot is greater than ANPP
in a drought plot was 0.71, and the probability that ANPP in a control
plot is lower than in an irrigation plot was 0.41. In other words, the
posterior distributions of
\(\beta_0(\text{control}) - \beta_0(\text{drought})\) and
\(\beta_0(\text{control}) - \beta_0(\text{drought})\) broadly overlapped
zero. There was also no evidence that the treatment effects became more
important over time because there was no directional trend in the yearly
treatment effects estimated from the ANPP-Treatment model (Fig. 3).

Community composition was similar among treatments. In no year did
community composition among treatments not overlap, and they were
equally dispersed in all years (Table 1; Fig. 4). Likewise, community
composition was remarkably stable over time, with no evidence of
divergence among treatments (Table 1; Fig. 4). Species' abundances and
ranks showed little deviation over the five-year experiment, regardless
of treatment (Fig. 5).

@@

Surprisingly, ANPP was not sensitive to either drought or irrigation
treatments, whether in the context of growing season precipitation (Fig.
2) or in the context of year since the experiment began (Fig. 3). This
is surprising because grasslands generally, and sagebrush steppe
specifically, are considered water-limited systems. Thus, our
expectation was that ANPP would be tightly linked to soil water
availability. ANPP was slightly lower in drought plots and slightly
higher in irrigation plots, but these differences were not statistically
important.

In the absence of community composition data, an obvious hypothesis to
explain the lack of ANPP response is that species reordered rapidly. In
so doing, newly abundant species could compensate for the loss of
ecosystem functioning of previously abundant species. We did not find
this to be the case, as specie composition was remarkably stable over
the course of the experiment (Fig. 3).

In combination, the lack of ecosystem and community response to drought
or irrigation shows that this sagebrush steppe ecosystem is resistant to
chronic alterations in water availability. There are three possible
explanations for the resistance we found. First, it could be that our
experiment simply was not long enough to induce responses. This may be
true for community responses, which can take decades \citep{Wilcox2016},
but the response of grassland ANPP to water manipulations is typically
immediate \citep[e.g.,][]{Hoover2014}.

Second, it could be that our manipulations were not large enough to
induce a response. That is, maybe a 50\% decrease/increase in any given
year is not abnormal give our sites historical range of variability
\citep{Knapp2017}. We cannot definitively rule out this possibility, but
we have reason to believe our manipulations \emph{should} have been
large enough. Using the methods described by \citet{Lemoine2016}, we
calculated the percent reduction and increase of mean growing season
precipitation neccesary to reach the 1\% and 99\% extremes of the
historical precipitation regime at our site. The 1\% quantile of
precipitation at our site is 110 mm, a 47\% reduction from the mean, and
the 99\% quantile is 414 mm, a 77\% increase from mean growing season
precipitation (Appendix 3). Thus, our drought treatment represented
extreme precipitation amounts, especially in years where ambient
precipitation was below average (Fig. 1A). The irrigation treatment may
have been too small, however.

Third, the ecosystem and community resistance to drought and deluge we
observed could be a real phenomenon in this system.

>>>>

\affil[1]{Department of Wildland Resources and the Ecology Center, Utah State
University, Logan, Utah 84322}
\affil[2]{Department of Ecology and Evolutionary Biology, University of
California, Los Angeles, Los Angeles, California 90095}

@@

% 
\begin{abstract}
Precipitation is predicted to become more variable in the western U.S.,
meaning years of above and below average precipitation will become more
common. Such periods of drought and deluge could become major drivers of
plant community dynamics and ecosystem functioning in water limited
grasslands. Here we report the results of a five-year experiment where
we used drought and irrigation treatments (50\% decrease/increase) to
see how a sagebrush steppe plant community in Idaho will respond to
future climate changes. Ecosystem functional response, defined as the
relationship between soil moisture and aboveground net primary
productivity, was suprisingly consistent across drought, control, and
irrigation treatments. Aboveground net primary productivity (ANPP)
responded positively to availabile soil moisture, but the response did
not vary across treatments. There was also no evidence that treatment
effects grew over time. The similarity of ecosystem functional response
across treatment was not due to compensatory shifts at the plant
community level, where species composition among treatments was similar
and remarkably stable over the five years. At least in the short-term,
ecosystem functional response and community composition in this
sagebrush steppe system is resistant to increases and decreases in
growing season precipitation.
% Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text.
\end{abstract}

@@

\noindent{}where \(\mu_{i(j(k))}\) is the deterministic prediction from
the regression model for observation \emph{i} for plot \emph{j}
associated with treatment \emph{k} in year \emph{t}, \(\beta_{0,j(k)}\)
is the intercept for plot \emph{j} associated with treatment \emph{k},
\(\beta_{1,j(k)}\) is the slope term for the effect of soil moisture for
plot \emph{j} associated with treatment \emph{k}, \(\gamma_t\) is the
intercept offset for year \emph{t}, and \(\sigma^2_k\) is the process
variance for treatment \emph{k}. Data include the standardized log(ANPP)
observations \(\left( y_{i(j(k(t)))} \right)\) and soil moisture
(\(x_i\)). Although we include observation subscript \emph{i} on the
\emph{x}s, observations within a treatment-year all share the same soil
moisture values.

@@

\noindent{}where \(\boldsymbol{\beta}_{j(k)}\) is the vector of
regression coefficients (intercept and slope) for plot \emph{j}
associated with treatment \emph{k}, \(\boldsymbol{\beta}_{k}\) is the
vector of coefficients for each treatment, and \(\boldsymbol{\beta}\) is
the vector of overall coefficients. The plot- and treatment-level
coefficients are drawn from multivariate normal distributions with
covariance matrix \(\boldsymbol{\Sigma}\). For the plot-level
coefficients, each treatment has its own variance-covariance matrix
(i.e., \(\boldsymbol{\Sigma}(k)\)). The overall coefficients are drawn
from a normal prior with mean 0 and standard deviation 1. The random
year effects (\(\boldsymbol{\gamma}\)) are drawn from a normal prior
with mean 0 and standard deviation \(\sigma_{\text{year}}\), which was
drawn from a weibull distribution. A full description of model is in
Appendix 2.

@@

Three of our five treatment years fell in years of below average
rainfall (Fig. 2A). Thus, those three years represent a lower magnitude
of absolute change in precipitation experienced by the treatments.
Averaged across treatments, ANPP varied from a minimum of 74.5 g
m\(^{-2}\) in 2014 to a maximum of 237.1 g m\(^{-2}\) in 2016 (Fig. 2C).
ANPP was slightly higher in irrigation plots and slightly lower in
drought plots (Fig. 2C), corresponding to estimated soil volumetric
water content (VWC) differences among treatments (Fig. 2B). Such
differences in soil VWC indicate our treatment infrastructure was
successful.

Cumulative growing season soil moisture had a positive effect on ANPP
(mean of \(\beta_{1}\) = 0.43; 80\% BCI = -0.05, 0.88; 95\% BCI = -0.36,
1.15) (Fig. 2D). Ecosystem functional response was similar among
treatments, with treatment level regressions between soil moisture and
log(ANPP) having similar intercepts (Fig. 3A) and slopes (Fig. 3B).
There was also no evidence that the treatment effects became more
important over time because there was no directional trend in the random
year effects (Fig. 4).

Community composition was similar among treatments. In no year did
community composition among treatments not overlap, and they were
equally dispersed in all years (Table 1; Fig. 5). Likewise, community
composition was remarkably stable over time, with no evidence of
divergence among treatments (Table 1; Fig. 5).

@@

Ecosystem functional response (i.e., the relationship between soil
moisture and ANPP) and community composition in this sagebrush steppe
ecosystem was consistent across drought, control, and irrigation
treatments (Fig. 2D, Fig. 3). In other words, functional responses
estimated from any precipitation level sampled apply to any other level
of precipitation, and the consistent functional response was not due to
species reordering and compensation. Thus, our results conform to the
first of our four predictions: neither ecosystem functional response nor
community composition changed with chronic alteration in water
availability (Fig. 1, top left).

This is surprising because grasslands generally, and sagebrush steppe
specifically, are considered water-limited systems. Thus, our naive
expecatations were that either (i) ecosystem functional response would
differ among treatments as species reach physiological thresholds and
resource limitations (Fig. 1, top right), or (ii) ecosystem functional
response stays the same due to species reordering and compensation (Fig.
1, bottom left).

In the absence of community composition data, an obvious hypothesis to
explain the consistent ecosystem functional response is that species
reordered rapidly. In so doing, newly abundant species could compensate
for the loss of ecosystem functioning of previously abundant species. We
did not find this to be the case, as species composition was remarkably
stable over the course of the experiment (Fig. 5).

In combination, the lack of ecosystem and community response to drought
or irrigation shows that this sagebrush steppe ecosystem is resistant to
chronic alterations in water availability. There are three possible
explanations for the resistance we found. First, it could be that our
experiment simply was not long enough to induce responses. This may be
true for community responses, which can take decades \citep{Wilcox2016}.

Second, it could be that our manipulations were not large enough to
induce a response. That is, maybe a 50\% decrease/increase in any given
year is not abnormal give our sites historical range of variability
\citep{Knapp2017}. We cannot definitively rule out this possibility, but
we have reason to believe our manipulations \emph{should} have been
large enough. Using the methods described by \citet{Lemoine2016}, we
calculated the percent reduction and increase of mean growing season
precipitation neccesary to reach the 1\% and 99\% extremes of the
historical precipitation regime at our site. The 1\% quantile of
precipitation at our site is 110 mm, a 47\% reduction from the mean, and
the 99\% quantile is 414 mm, a 77\% increase from mean growing season
precipitation (Appendix 3). Thus, our drought treatment represented
extreme precipitation amounts, especially in years where ambient
precipitation was below average (Fig. 2A). The irrigation treatment may
have been too small, however.

Third, the ecosystem and community resistance to drought and deluge we
observed could be a real phenomenon in this system, at least over the
five year time span we observed. The perennial plants in this cold
desert ecosystem are tolerant to drought conditions \citep[A.R.
Kleinhesselink, unpublished data]{Bazzaz1979, Franks2011}, meaning a
consistent ecosystem functional response at low soil moisture is
perfectly reasonable. For example, many of the perennial grasses in our
focal ecosystem avoid drought stress by growing early in the growing
season (A.R. Kleinhesselink, personal observation). Likewise, the
dominant shrub in our focal ecosystem, \emph{Artemisia tripartita}, has
access to water deep in the soil profile thanks to a deep root system
\citep{Germino2014}. As for the irrigation treatments, the lack of a
saturating response (e.g., a lower regression slope) suggests water
availability remained a limiting factor at the levels of precipitation
we were able to manipulate.

In conclusion, our results suggest that five years of \(\pm\) 50\%
ambient precipitation is not enough to induce a shift in ecosystem
functional response in a sagebrush steppe. This is likely because
component species are tolerant to drought conditions and water remains
co-limiting at higher precipitation levels. Longer time series of
chronic precipitation alteration may reveal plant community shifts that
we did not observe \citep[e.g.,][]{Wilcox2016}. Our results suggest
compositional shifts would have the largest impact at high rainfall
because the current community maintained consistent ecosystem functional
response at very low water availability.
